StateSpaceModels.jl: a Julia Package for Time-Series Analysis in a State-Space Framework

### 2020-02-20 paper/paper.tex
Update paper (#151)

The contributions brought by these packages notwithstanding, there are currently no state-space model packages aimed at time-series analysis in Julia. Furthermore, some have not been updated for a long time and are not functional in Julia 1.0. Thus, the main objectives of \texttt{StateSpaceModels.jl} \cite{statespacemodels} are:

@@

The local level model consists of a stochastic level component that is defined by a random walk:
%
\begin{align}
    y_{t} &= \mu_{t} + \varepsilon_{t}, \quad \varepsilon_{t} \sim N(0, \sigma^{2}_\varepsilon), \label{eq:locallevel1} \\
    \mu_{t+1} &= \mu_{t} + \xi_{t}, \quad \xi_{t} \sim N(0, \sigma^{2}_\xi). \label{eq:locallevel2}
\end{align}
%
A local level \texttt{StateSpaceModel} can be created using:
%
\begin{lstlisting}[language = Julia]
model = local_level(y)
\end{lstlisting}
%
\noindent where \texttt{y} is an observation vector in the univariate case or a matrix in column-wise fashion in the multivariate case, i.e., each column representing a variable and each line representing a time period.
%

@@

The linear trend model consists of a local level model with the addition of a stochastic slope term, as seen in Eq. \eqref{eq:lineartrend1}--\eqref{eq:lineartrend3}. A linear trend \texttt{StateSpaceModel} can be created using:
%
\begin{lstlisting}[language = Julia]
model = linear_trend(y)
\end{lstlisting}

@@

The basic structural model consists of three stochastic components: trend, slope, and seasonality, defined in the following manner:
%
\begin{align}
    y_{t} &= \mu_{t} + \gamma_{t} + \varepsilon_{t}, \hspace{0.09\columnwidth} \varepsilon_{t} \sim N(0, \sigma^{2}_\varepsilon), \label{eq:structural1} \\
    \mu_{t+1} &= \mu_{t} + \nu_{t} + \xi_{t}, \hspace{0.095\columnwidth} \xi_{t} \sim N(0, \sigma^{2}_\xi), \label{eq:structural2} \\
    \nu_{t+1} &= \nu_{t} + \zeta_{t}, \hspace{0.18\columnwidth} \zeta_{t} \sim N(0, \sigma^{2}_\zeta), \label{eq:structural3} \\
    \gamma_{t+1} &= \sum_{j=1}^{s-1} \gamma_{t+1-j} + \omega_{t}, \quad \omega_{t} \sim N(0, \sigma^{2}_\omega). \label{eq:structural4}
\end{align}
%
In this model, there is a stochastic trend that contains a slope component, similarly to the linear trend model. Furthermore, a stochastic seasonal component is modeled so that its sum over the seasonality period $s$ is equal to zero except for a noise factor. A basic structural \texttt{StateSpaceModel} can be created using:
%
\begin{lstlisting}[language = Julia]
model = structural(y, s)
\end{lstlisting}

@@

The filtering step derives the predictive and filtered states as well as their covariance matrices at every time period via the Kalman filter. The predictive state and its variance are given by
%
\begin{align}
    a_{t+1} &= \mathbb{E}[\alpha_{t+1} | Y_{t}], \\
    P_{t+1} &= \mathbb{V}[\alpha_{t+1} | Y_{t}],
\end{align}
%
while the filtered state and its variance are given by
%
\begin{align}
    a_{t|t} &= \mathbb{E}[\alpha_{t} | Y_{t}], \\
    P_{t|t} &= \mathbb{V}[\alpha_{t} | Y_{t}],
\end{align}
%
where $Y_{t}$ denotes the set of time periods $y_{1}, \dots, y_{t}$.

@@

The Kalman filter is responsible for deriving estimates for the predictive state and for its covariance matrix at each time period, but it is dependent on the values of the noise covariances $H$ and $Q$. Thus, these constant parameters need to be estimated via maximum likelihood \cite{casella2002statistical}. The log-likelihood function is given by
%
\begin{align}
    \ell(Y_{n}) = -\frac{np}{2} \log 2 \pi - \frac{1}{2} \sum_{t=1}^{n} (\log |F_{t}| + v_{t}^{\top} F_{t}^{-1} v_{t}),
\end{align}
%
which depends on the innovations and their covariance matrix at each time period. Therefore, at each optimization iteration, the Kalman filter needs to be executed.

@@

As a way of evaluating the model specification and estimation, the package contains statistical diagnostics that can be conducted over the residuals. The available diagnostics are the Jarque-Bera normality test \cite{jarque1980efficient}, the Ljung-Box independence test \cite{box1970distribution}, and an homoscedasticity test \cite{durbin2012time}. The diagnostics can be run through function \texttt{diagnostics}.

@@

As a first example, let us use the classical monthly airline passengers time series. In order to avoid multiplicative effects, we use the well-known approach of taking the log of the series. Fig. \ref{fig:log_airline_forecast} shows the log-airline passengers time series. We can estimate a structural \texttt{StateSpaceModel} as follows.
%
\begin{lstlisting}[language = Julia]
# Load the AirPassengers dataset
AP = CSV.read("AirPassengers.csv")
# Take the log of the series
logAP = log.(Vector{Float64}(AP[:Passengers]))
# Specify the state-space model
model = structural(logAP, 12)
# Estimate the state-space model
ss = statespace(model)
\end{lstlisting}

By estimating a structural model, we can analyzed the individual components of the series, such as trend and seasonality. These components are presented in Fig. \ref{fig:log_airline_components} and represent dimensions 1 and 3 of the smoothed state, respectively.

>>>>

The contributions brought by these packages notwithstanding, there are currently no state-space model packages aimed at time-series analysis in Julia. Thus, the main objectives of \texttt{StateSpaceModels.jl} \cite{statespacemodels} are:

@@

The local level model consists of a stochastic level component that is defined by a random walk:
%
\begin{align}
    y_{t} &= \mu_{t} + \varepsilon_{t}, \quad \varepsilon_{t} \sim N(0, \sigma^{2}_\varepsilon), \label{eq:locallevel1} \\
    \mu_{t+1} &= \mu_{t} + \xi_{t}, \quad \xi_{t} \sim N(0, \sigma^{2}_\xi). \label{eq:locallevel2}
\end{align}
%
A local level \texttt{StateSpaceModel} can be created using:
%
\begin{lstlisting}[language = Julia]
model = local_level(y)
\end{lstlisting}
%
\noindent where \texttt{y} is an observation vector in the univariate case or a matrix in column-wise fashion in the multivariate case, i.e., each column representing a variable and each line representing a time period.

While very simple, the local level model can be successfully applied to a wide array of problems wherein one wants to estimate some magnitude that varies over time. For instance, in \cite{koopman2000stamp}, an illustrative example utilizes water inflow data from the river Nile.
%

@@

The linear trend model consists of a local level model with the addition of a stochastic slope term, as seen in Eq. \eqref{eq:lineartrend1}--\eqref{eq:lineartrend3}. Note that it differs from the local level model due to the inclusion of the slope $\nu_{t}$ in \eqref{eq:lineartrend2}. A linear trend \texttt{StateSpaceModel} can be created using:
%
\begin{lstlisting}[language = Julia]
model = linear_trend(y)
\end{lstlisting}

@@

The basic structural model consists of three stochastic components: trend, slope, and seasonality, defined in the following manner:
%
\begin{align}
    y_{t} &= \mu_{t} + \gamma_{t} + \varepsilon_{t}, \hspace{0.09\columnwidth} \varepsilon_{t} \sim N(0, \sigma^{2}_\varepsilon), \label{eq:structural1} \\
    \mu_{t+1} &= \mu_{t} + \nu_{t} + \xi_{t}, \hspace{0.095\columnwidth} \xi_{t} \sim N(0, \sigma^{2}_\xi), \label{eq:structural2} \\
    \nu_{t+1} &= \nu_{t} + \zeta_{t}, \hspace{0.18\columnwidth} \zeta_{t} \sim N(0, \sigma^{2}_\zeta), \label{eq:structural3} \\
    \gamma_{t+1} &= \sum_{j=1}^{s-1} \gamma_{t+1-j} + \omega_{t}, \quad \omega_{t} \sim N(0, \sigma^{2}_\omega). \label{eq:structural4}
\end{align}
%
In this model, there is a stochastic trend that contains a slope component, exactly as the linear trend model. Furthermore, a stochastic seasonal component is modeled such that its sum over the seasonality period $s$ is equal to zero except for a noise factor. A basic structural \texttt{StateSpaceModel} can be created using:
%
\begin{lstlisting}[language = Julia]
model = structural(y, s)
\end{lstlisting}

The idea of the seasonal component is to account for effects that have a periodic nature, such as monthly temperature over each year.

@@

The filtering step computes the predictive and filtered states as well as their covariance matrices at every time period via the Kalman filter. The predictive state and its variance are given by
%
\begin{align}
    a_{t+1} &= \mathbb{E}[\alpha_{t+1} | Y_{t}], \\
    P_{t+1} &= \mathbb{V}[\alpha_{t+1} | Y_{t}],
\end{align}
%
while the filtered state and its variance are given by
%
\begin{align}
    a_{t|t} &= \mathbb{E}[\alpha_{t} | Y_{t}], \\
    P_{t|t} &= \mathbb{V}[\alpha_{t} | Y_{t}],
\end{align}
%
where $Y_{t}$ denotes the set of time periods $y_{1}, \dots, y_{t}$.

@@

The Kalman filter is responsible for deriving estimates for the predictive state and for its covariance matrix at each time period, but it is dependent on the values of the noise covariances $H$ and $Q$. Thus, these constant parameters need to be estimated via maximum likelihood \cite{casella2002statistical}. The log-likelihood function is given by
%
\begin{align}
    \ell(Y_{n}) = -\frac{np}{2} \log 2 \pi - \frac{1}{2} \sum_{t=1}^{n} (\log |F_{t}| + v_{t}^{\top} F_{t}^{-1} v_{t}),
\end{align}
%
which depends on the innovations and their covariance matrix at each time period. Therefore, at each iteration of the optimization, the Kalman filter needs to be executed to compute the objective function.

@@

As a way of evaluating the model specification and estimation, the package contains basic statistical diagnostics that can be conducted over the residuals. The available diagnostics are the Jarque-Bera normality test \cite{jarque1980efficient}, the Ljung-Box independence test \cite{box1970distribution}, and an homoscedasticity test \cite{durbin2012time}. The diagnostics can be run through function \texttt{diagnostics}.

@@

As a first example, let us use the classical monthly airline passengers time series. In order to avoid multiplicative effects, we use the well-known approach of taking the log of the series \cite{lutkepohl2012role}. Fig. \ref{fig:log_airline_forecast} shows the log-airline passengers time series. We can estimate a structural \texttt{StateSpaceModel} as follows.
%
\begin{lstlisting}[language = Julia]
# Load the AirPassengers dataset
AP = CSV.read("AirPassengers.csv")
# Take the log of the series
logAP = log.(Vector{Float64}(AP[:Passengers]))
# Specify the state-space model
model = structural(logAP, 12)
# Estimate the state-space model
ss = statespace(model)
\end{lstlisting}

By estimating a structural model, we can analyze the individual components of the series, such as trend and seasonality. These components are presented in Fig. \ref{fig:log_airline_components} and represent dimensions 1 and 3 of the smoothed state, respectively.
